Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-03-03 sources/bmc_template_submission/bmc_article.tex
adding verificaion test

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases,
with the main goal of building on published alignments and extending existing
phylogenetic inferences with more data and taxa.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). The tree sampling 41 species is also available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

@@

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases,
with the main goal of building upon published alignments to extend existing
phylogenetic inferences with more data and taxa.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). The tree samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the tree from \cite{yao2020phylogeny} is the best sampled phylogeny yet available for
the hollies.

@@

\subsection*{Verification test}

We pruned 9 out of the 41 original tips of the Gottlieb et al. tree, corresponding
to a 20\% trim, excluding the outgroups. We then performed a Physcraper run to test
if we would recover the dropped tips.
We successfully recovered 6 out of 9 pruned tips in the updated tree.
Closer examination of results revealed that the 3 missing tips were correctly retrieved
with BLAST along with the other 6 tips that were included in the updated tree. Examining
the ITS sequences belonging to the tips missed by the test, by following the GenBank
accession numbers reported in the original publication, showed that these sequences
contain a 100 bp long gap of unidentified nucleotides (Ns). These Ns were absent
in the 3 sequences published in the original alignment.
